### 2016-09-13 Chapters/QuantumMechanics.tex
First draft sent to L.F. and B.M.
Multipliers equations for CC2 and CC3 added

@@

Introducing the \gls{MP} partitioning of the Hamiltonian leads to
further insight:
\begin{alignat}{2}
 H = F + \Phi = \sum_r\epsilon_r\cons{r}\anni{r} + (g- V_\mathrm{HF}),
 \quad&
 V_\mathrm{HF} = \sum_{i}[g_{rsii} - g_{riis}]\cons{r}\anni{s}
\end{alignat}
where $F$ if the Fock matrix and $\Phi$ is the fluctuation potential.
Given our initial assumption on the reference function, the Fock
operator is diagonal and expressed in terms of spin-orbital energies and
number operators.
Its similarity transformation truncates after the second term and has a
relatively compact form:
\begin{equation}
  \overline{F} = F + \sum_{i}^4\sum_{\mu_i}\tamp{\mu_i}\cluster{\mu_i}\denom{\mu_i}
\end{equation}
where $\denom{\mu_i}$ is the difference in orbital energies between the
virtual and occupied spin-orbitals of excitation $i$.
For the fluctuation potential the similarity transformation truncates
after the fifth term, including up to four-fold nested commutators.

As formulated so far, the \acrshort{CC} method is not variational.
It is however possible to introduce a fully variational Lagrangian that
leads to the same amplitudes equations.\autocite{Helgaker1982-xl, Arponen1983-qt,
Helgaker1988-to, Helgaker1989-wl, Koch1990-vr, Kvaal2013-jr}
Let us call $\mathcal{E}(\vect{\eta})$ the definition of the energy
for the given quantum chemical method.
The \emph{amplitudes} $\vect{\eta}$ are determined by a set of
stationarity conditions $\mat{\Omega}(\vect{\eta}) = 0$.
Direct minimization of $\mathcal{E}(\vect{\eta})$ is thus a
\emph{constrained} optimization.
Introducing a set of Lagrange \emph{multipliers} $\bar{\vect{\eta}}$
lets us recast the problem as an unconstrained minimization of the
Lagrangian:
\begin{equation}
  \mathcal{L}(\vect{\eta}, \bar{\vect{\eta}}) =
  \mathcal{E}(\vect{\eta}) +
  \scalprod[W]{\bar{\vect{\eta}}}{\mat{\Omega}(\vect{\eta})}
\end{equation}
where the governing equations are obtained by differentiation with
respect to both sets of parameters:
\begin{subequations}
  \begin{align}
    \pderiv{\mathcal{L}(\vect{\eta}, \bar{\vect{\eta}})}{\bar{\vect{\eta}}}
    &= \mat{\Omega}(\vect{\eta}) = 0 \\
    \pderiv{\mathcal{L}(\vect{\eta}, \bar{\vect{\eta}})}{\vect{\eta}}
    &= \pderiv{\mathcal{E}(\vect{\eta})}{\vect{\eta}} +
    \scalprod[W]{\bar{\vect{\eta}}}{\pderiv{\mat{\Omega(\vect{\eta})}}{\vect{\eta}}}
    = 0
  \end{align}
\end{subequations}
The \acrshort*{CC} Lagrangian can thus be expressed as:
\begin{equation}\label{eq:cc-lagrangian}
  \begin{aligned}
    \mathcal{L}^\mathcal{M}(\tamp{}, \tbar{})
    &=
    \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
    + \sum_{i=1}^{\mathcal{M}}\sum_{\mu_i}
    \tbar{\mu_i}\braket{\mu_i | \overline{H} | \mathrm{HF}} \\
    &=
  E_0
  + \sum_{i=1}^{\mathcal{M}}\tbar{\mu_i}\denom{\mu_i}\tamp{\mu_i}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}}
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{\Phi} | \mathrm{HF}}
  \end{aligned}
\end{equation}
where $E_0 = \sum_{i}\epsilon_i$ is the sum of the one-electron orbital
energies and the shorthand notation for the Lagrangian multiplier state (left
\acrshort*{CC} state) was introduced:
\begin{equation}
  \bra{\tbar{j}} = \sum_{\mu_j}\tbar{\mu_j}\bra{\mu_j}.
\end{equation}
Differentiation of the Lagrangian with respect to the multipliers
correctly yields the amplitudes equations, while differentiation with
respect to the amplitudes leads to the governing equations for the
multipliers:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_i}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_i}\tamp{\mu_i} + \braket{\mu_i | \overline{\Phi} | \mathrm{HF}}
             = 0 \label{eq:cc-amplitudes}\\
   \tbarEq{\mu_i}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_i}\tbar{\mu_i} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}} +
    \sum_{j}^{\mathcal{M}}\braket{\tbar{j} |
    \commutator{\overline{\Phi}}{\cluster{\mu_i}} | \mathrm{HF}}
             = 0 \label{eq:cc-multipliers}
  \end{align}
\end{subequations}
While the multipliers are not needed for the evaluation of the
\acrshort{CC} energy, their calculation is mandatory when expectation values
and molecular properties in general are sought.
The \acrshort{CC} expectation values are formed using the left and
right \acrshort{CC} states and we introduce the following notation:
\begin{equation}
  O(\tamp{}, \tbar{}) = \braket{\mathrm{HF} | \overline{O} | \mathrm{HF} }
  + \sum_{i=1}^{\mathcal{M}}\braket{\tbar{i} | \overline{O} | \mathrm{HF} }
\end{equation}
When required by the context, we will explicitly write at which
truncation level in the cluster operator the expectation was evaluated:
$O(\tamp{}, \tbar{})_{\mathrm{CC}\mathcal{M}}$.

@@

It is clear from the stationarity conditions that $\tamp{\mu_i}^{[0]} = 0, \,\forall i$ and
$\tbar{\mu_i}^{[0]} = 0, \, \forall i$.
Thus cluster operators can be of order 1 and higher. We introduce the
following notation for the $i$-th order cluster operators and
multipliers states:
\begin{alignat}{2}
  T_p^{[i]} = \sum_{\mu_p}\tamp{\mu_p}^{[i]}\cluster{\mu_p},
  \quad&
  \bra{\tbar{p}^{[i]}} = \sum_{\mu_p}\tbar{\mu_p}^{[i]}\bra{\mu_p}
\end{alignat}

To first order, one has:
\begin{equation}\label{eq:1st-order}
  \begin{aligned}
  \denom{\mu_2}\tamp{\mu_2}^{[1]}
  &= - \braket{\mu_2 | \Phi | \mathrm{HF}}
  = - \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_2}} |
  \mathrm{HF}} \\
  &= \denom{\mu_2}\tbar{\mu_2}^{[1]}
  \end{aligned}
\end{equation}
which shows that singles, due to Brillouin's condition, and triples
amplitudes are of second and higher order.
To second order, one obtains:
\begin{equation}\label{eq:2nd-order}
  \begin{aligned}
  \denom{\mu_i}\tamp{\mu_i}^{[2]}
  &= - \braket{\mu_i | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_i}} |
  \mathrm{HF}} \\
  &= \denom{\mu_i}\tbar{\mu_i}^{[2]},\quad i = 1, 2, 3.
  \end{aligned}
\end{equation}
According to the $2n+1$ and $2n+2$ rules, energy corrections up to fifth
order can be obtained from amplitudes and multipliers up to second
order. Since the triples first appear to second order, the \gls{CCSD}
energy is correct to third order in the fluctuation potential.
The triples contributions to the fourth and fifth order corrections to
the energy are given as:
\begin{subequations}
  \begin{align}
    E_T^{[4]} &=
 \braket{\tbar{2}^{[1]}| \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
 \label{eq:4th-order-triples} \\
    E_T^{[5]} &=
    \braket{\tbar{1}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{2}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
  &+ \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_2^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
    &+
    \frac{1}{2}\braket{\tbar{3}^{[2]} |
    \commutator{\commutator{\Phi}{T_2^{[1]}}}{T_2^{[1]}}
    | \mathrm{HF}}.
 \label{eq:5th-order-triples}
  \end{align}
\end{subequations}
These expressions form the basis for the development of reduced-scaling
\acrshort*{CC} schemes were the double and triple excitation are treated
approximately. These are the subject of the next Section.

@@

In the iterative schemes, one approximates the amplitude equations for
the higher excitation to be correct up to a given order in perturbation
theory.
For the development of iterative schemes, we assume that $T_1 = O(0)$
and use $T_1$-transformed operators:
\begin{equation}
  \check{O} = \expo{-T_1}O\expo{T_1}.
\end{equation}
Since $T_1$ is a one-electron operator, the $T_1$-transformation does
not affect the particle rank of the electronic
Hamiltonian.\autocite{Koch1994-vr, Helgaker2000-tz}
The assumption that $T_1$ is of zeroth order is contrary to the usual
\acrshort*{MP} approach but highlights the role of the singles
amplitudes as orbital optimization parameters.
In the \gls{CC2} method, the amplitudes equations are:\autocite{Christiansen1995-vy, Helgaker2000-tz}
\begin{subequations}
  \begin{align}
    \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 | \check{\Phi} +
    \commutator{\check{\Phi}}{Q_2} | \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tamp{\mu_2} &+ \braket{\mu_2 | \check{\Phi} | \mathrm{HF}} = 0
  \end{align}
\end{subequations}
and are obtained from the \acrshort{CCSD} equations by retaining only
first order terms in the doubles equations.
The approximate doubles cluster operator was introduced:
\begin{equation}
  Q_2 = - \sum_{\mu_2}
  \denom{\mu_2}^{-1}\braket{\mu_2 | \check{\Phi} | \mathrm{HF}}\cluster{\mu_2}
\end{equation}
and we shall refer to it as quasi-$T_2$.
The \acrshort*{CC2} Lagrangian is constructed based on the approximate
equations obtained for the singles and
doubles:\autocite{Christiansen1995-vy, Helgaker2000-tz}
\begin{equation}\label{eq:cc2-lagrangian}
  \begin{aligned}
  \mathcal{L}_\mathrm{CC2}(\tamp{}, \tbar{})
  &=
  E_0
  + \sum_{i=1}^{\mathcal{2}}\tbar{\mu_i}\denom{\mu_i}\tamp{\mu_i}
  + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{Q_2} | \mathrm{HF}} \\
  &+ \sum_{\mu_1}\tbar{\mu_1}\braket{\mu_1 |
  \check{\Phi} + \commutator{\check{\Phi}}{Q_2}
  | \mathrm{HF}} \\
  &+ \sum_{\mu_2}\tbar{\mu_2}\braket{\mu_2 | \check{\Phi} | \mathrm{HF}}
  \end{aligned}
\end{equation}
The \acrshort*{CC2} methods brings the scaling of the \acrshort*{CCSD}
method down to $N^5$ in the iterative step.

The \gls{CC3} model is constructed analogously, the amplitudes
equations:\autocite{Christiansen1995-pv, Koch1997-nm, Helgaker2000-tz,
Smith2005-an}
\begin{subequations}
  \begin{align}
    \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 |
    \check{\Phi} + \commutator{\check{\Phi}}{T_2}
    + \commutator{\check{\Phi}}{Q_3}
    | \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tamp{\mu_2} &+ \braket{\mu_2 |
    \check{\Phi} + \commutator{\check{\Phi}}{T_2}
    + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2}
    + \commutator{\check{\Phi}}{Q_3}
    | \mathrm{HF}} = 0 \\
    \denom{\mu_3}\tamp{\mu_3} &+ \braket{\mu_3 |
    \commutator{\check{\Phi}}{T_2}
    | \mathrm{HF}} = 0
  \end{align}
\end{subequations}
are obtained from the \acrshort*{CCSDT} equations by retaining only
second order terms in the triples equation.
The approximate triples cluster operator was introduced:
\begin{equation}
  Q_3 = - \sum_{\mu_3}
  \denom{\mu_3}^{-1}\braket{\mu_3 | \commutator{\check{\Phi}}{T_2} | \mathrm{HF}}\cluster{\mu_3}
\end{equation}
and we shall refer to it as quasi-$T_3$.
The \acrshort*{CC3} Lagrangian is constructed based on the approximate
equations obtained for the singles, doubles and
triples:\autocite{Christiansen1995-pv, Koch1997-nm, Helgaker2000-tz, Smith2005-an}
\begin{equation}\label{eq:cc3-lagrangian}
  \begin{aligned}
  \mathcal{L}_\mathrm{CC3}(\tamp{}, \tbar{})
  &=
  E_0
  + \sum_{i=1}^{\mathcal{3}}\tbar{\mu_i}\denom{\mu_i}\tamp{\mu_i}
  + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}} \\
  &+ \sum_{\mu_1}\tbar{\mu_1}\braket{\mu_1 |
  \check{\Phi} + \commutator{\check{\Phi}}{T_2} + \commutator{\check{\Phi}}{Q_3}
  | \mathrm{HF}} \\
  &+ \sum_{\mu_2}\tbar{\mu_2}\braket{\mu_2 |
  \check{\Phi} + \commutator{\check{\Phi}}{T_2}
  + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{Q_3}
  | \mathrm{HF}} \\
  &+ \sum_{\mu_3}\tbar{\mu_3}\braket{\mu_3 | \commutator{\check{\Phi}}{T_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
The \acrshort*{CC3} methods brings the scaling of the \acrshort*{CCSDT}
method down to $N^7$ in the iterative step.

>>>>

\begin{epigraphs}
\setlength\epigraphwidth{7.5cm}
\qitem{
plays with the bigness of his littleness

--- electrons deify one razorblade

into a mountainrange; lenses extend

unwish through curving wherewhen till unwish

returns on its unself.
    }{
    --- \textsc{E. E. Cummings}}
\setlength\epigraphwidth{7cm}
\end{epigraphs}

@@

Introducing the \gls{MP} partitioning of the Hamiltonian leads to
further insight:
\begin{alignat}{2}
 H = F + \Phi = \sum_r\epsilon_r\cons{r}\anni{r} + (g- V_\mathrm{HF}),
 \quad&
 V_\mathrm{HF} = \sum_{i}[g_{rsii} - g_{riis}]\cons{r}\anni{s}
\end{alignat}
where $F$ if the Fock matrix and $\Phi$ is the fluctuation potential.
Given our initial assumption on the reference function, the Fock
operator is diagonal and expressed in terms of spin-orbital energies and
number operators.
Its similarity transformation truncates after the second term and has a
relatively compact form:
\begin{equation}
  \overline{F} = F + \sum_{u=1}^\mathcal{M}\sum_{\mu_u}\denom{\mu_u}\tamp{\mu_u}\cluster{\mu_u}
\end{equation}
where $\denom{\mu_u}$ is the difference in orbital energies between the
virtual and occupied spin-orbitals of excitation $u$.
For the fluctuation potential the similarity transformation truncates
after the fifth term, including up to four-fold nested commutators.

As formulated so far, the \acrshort{CC} method is not variational.
It is however possible to introduce a fully variational Lagrangian that
leads to the same amplitudes equations.\autocite{Helgaker1982-xl, Arponen1983-qt,
Helgaker1988-to, Helgaker1989-wl, Koch1990-vr, Kvaal2013-jr}
Let us call $\mathcal{E}(\vect{\eta})$ the definition of the energy
for the given quantum chemical method.
The \emph{amplitudes} $\vect{\eta}$ are determined by a set of
stationarity conditions $\mat{\Omega}(\vect{\eta}) = 0$.
Direct minimization of $\mathcal{E}(\vect{\eta})$ is thus a
\emph{constrained} optimization.
Introducing a set of Lagrange \emph{multipliers} $\bar{\vect{\eta}}$
lets us recast the problem as an unconstrained minimization of the
Lagrangian:
\begin{equation}
  \mathcal{L}(\vect{\eta}, \bar{\vect{\eta}}) =
  \mathcal{E}(\vect{\eta}) +
  \scalprod[W]{\bar{\vect{\eta}}}{\mat{\Omega}(\vect{\eta})}
\end{equation}
where the governing equations are obtained by differentiation with
respect to both sets of parameters:
\begin{subequations}
  \begin{align}
    \pderiv{\mathcal{L}(\vect{\eta}, \bar{\vect{\eta}})}{\bar{\vect{\eta}}}
    &= \mat{\Omega}(\vect{\eta}) = 0 \\
    \pderiv{\mathcal{L}(\vect{\eta}, \bar{\vect{\eta}})}{\vect{\eta}}
    &= \pderiv{\mathcal{E}(\vect{\eta})}{\vect{\eta}} +
    \scalprod[W]{\bar{\vect{\eta}}}{\pderiv{\mat{\Omega(\vect{\eta})}}{\vect{\eta}}}
    = 0
  \end{align}
\end{subequations}
The \acrshort*{CC} Lagrangian can thus be expressed as:
\begin{equation}\label{eq:cc-lagrangian}
  \begin{aligned}
    \mathcal{L}(\tamp{}, \tbar{})_\mathcal{M}
    &=
    \braket{\mathrm{HF} | \overline{H} | \mathrm{HF}}
    + \sum_{u=1}^{\mathcal{M}}\sum_{\mu_u}
    \tbar{\mu_u}\braket{\mu_u | \overline{H} | \mathrm{HF}} \\
    &=
  E_0
  + \sum_{u=1}^{\mathcal{M}}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
  + \braket{\mathrm{HF} | \overline{\Phi} | \mathrm{HF}}
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{\Phi} | \mathrm{HF}}
  \end{aligned}
\end{equation}
where $E_0 = \sum_{i}\epsilon_i$ is the sum of the one-electron orbital
energies and the shorthand notation for the Lagrangian multiplier state (left
\acrshort*{CC} state) was introduced:
\begin{equation}
  \bra{\tbar{u}} = \sum_{\mu_u}\tbar{\mu_u}\bra{\mu_u}.
\end{equation}
Differentiation of the Lagrangian with respect to the multipliers
correctly yields the amplitudes equations, while differentiation with
respect to the amplitudes leads to the governing equations for the
multipliers:
\begin{subequations}
  \begin{align}
   \tampEq{\mu_q}(\tamp{}, \tbar{}, \p)  &=
   \denom{\mu_q}\tamp{\mu_q} + \braket{\mu_q | \overline{\Phi} | \mathrm{HF}}
             = 0 \label{eq:cc-amplitudes}\\
   \tbarEq{\mu_q}(\tamp{}, \tbar{}, \p)
    &=
    \denom{\mu_q}\tbar{\mu_q} +
    \braket{\mathrm{HF} | \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}} +
    \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} |
    \commutator{\overline{\Phi}}{\cluster{\mu_q}} | \mathrm{HF}}
             = 0 \label{eq:cc-multipliers}
  \end{align}
\end{subequations}
While the multipliers are not needed for the evaluation of the
\acrshort{CC} energy, their calculation is mandatory when expectation values
and molecular properties in general are sought.
The \acrshort{CC} expectation values are formed using the left and
right \acrshort{CC} states and we introduce the following notation:
\begin{equation}
  O(\tamp{}, \tbar{})_\mathcal{M} = \braket{\mathrm{HF} | \overline{O} | \mathrm{HF} }
  + \sum_{u=1}^{\mathcal{M}}\braket{\tbar{u} | \overline{O} | \mathrm{HF} }
\end{equation}

@@

It is clear from the stationarity conditions that $\tamp{\mu_u}^{[0]} = 0, \,\forall u$ and
$\tbar{\mu_u}^{[0]} = 0, \, \forall u$.
Thus cluster operators can be of order 1 and higher. We introduce the
following notation for the $i$-th order cluster operators and
multipliers states:
\begin{alignat}{2}
  T_u^{[i]} = \sum_{\mu_u}\tamp{\mu_u}^{[i]}\cluster{\mu_u},
  \quad&
  \bra{\tbar{u}^{[i]}} = \sum_{\mu_u}\tbar{\mu_u}^{[i]}\bra{\mu_u}
\end{alignat}

To first order, one has:
\begin{equation}\label{eq:1st-order}
  \denom{\mu_2}\tamp{\mu_2}^{[1]}
  = - \braket{\mu_2 | \Phi | \mathrm{HF}}
  = - \braket{\mathrm{HF} | \commutator{\Phi}{\cluster{\mu_2}} |
  \mathrm{HF}}
  = \denom{\mu_2}\tbar{\mu_2}^{[1]}
\end{equation}
which shows that singles, due to Brillouin's condition, and triples
amplitudes are of second and higher order.
For second order singles, doubles and triples one obtains:
\begin{equation}\label{eq:2nd-order}
  \denom{\mu_q}\tamp{\mu_q}^{[2]}
  = - \braket{\mu_q | \commutator{\Phi}{T_2^{[1]}} | \mathrm{HF}}
  = - \braket{\tbar{2}^{[1]} | \commutator{\Phi}{\cluster{\mu_q}} |
  \mathrm{HF}}
  = \denom{\mu_q}\tbar{\mu_q}^{[2]}
\end{equation}
According to the $2n+1$ and $2n+2$ rules, energy corrections up to fifth
order can be obtained from amplitudes and multipliers up to second
order. Since the triples first appear to second order, the \gls{CCSD}
energy is correct to third order in the fluctuation potential.
The triples contributions to the fourth and fifth order corrections to
the energy are given as:
\begin{subequations}
  \begin{align}
    E_T^{[4]} &=
 \braket{\tbar{2}^{[1]}| \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
 \label{eq:4th-order-triples} \\
    E_T^{[5]} &=
    \braket{\tbar{1}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{2}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
  &+ \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_2^{[2]}} | \mathrm{HF}}
  + \braket{\tbar{3}^{[2]} | \commutator{\Phi}{T_3^{[2]}} | \mathrm{HF}}
  \nonumber \\
    &+
    \frac{1}{2}\braket{\tbar{3}^{[2]} |
    \commutator{\commutator{\Phi}{T_2^{[1]}}}{T_2^{[1]}}
    | \mathrm{HF}}.
 \label{eq:5th-order-triples}
  \end{align}
\end{subequations}
These expressions form the basis for the development of reduced-scaling
\acrshort*{CC} schemes were the double and triple excitation are treated
approximately. These are the subject of the next Section.

@@

In the iterative schemes, one approximates the amplitude equations for
the higher excitation to be correct up to a given order in perturbation
theory.
For the development of iterative schemes, we assume that $T_1 = O(0)$
and use $T_1$-transformed operators:
\begin{equation}
  \check{O} = \expo{-T_1}O\expo{T_1}.
\end{equation}
Since $T_1$ is a one-electron operator, the $T_1$-transformation does
not affect the particle rank of the electronic
Hamiltonian.\autocite{Koch1994-vr, Helgaker2000-tz}
The assumption that $T_1$ is of zeroth order is contrary to the usual
\acrshort*{MP} approach but highlights the role of the singles
amplitudes as orbital optimization parameters.
In the \gls{CC2} method, the amplitudes equations are:\autocite{Christiansen1995-vy, Helgaker2000-tz}
\begin{subequations}\label{eq:cc2-amplitudes}
  \begin{align}
    \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 | \check{\Phi} +
    \commutator{\check{\Phi}}{Q_2} | \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tamp{\mu_2} &+ \braket{\mu_2 | \check{\Phi} | \mathrm{HF}} = 0
  \end{align}
\end{subequations}
and are obtained from the \acrshort{CCSD} equations by retaining only
first order terms in the doubles equations.
The approximate doubles cluster operator was introduced:
\begin{equation}
  Q_2 = - \sum_{\mu_2}
  \denom{\mu_2}^{-1}\braket{\mu_2 | \check{\Phi} | \mathrm{HF}}\cluster{\mu_2}
\end{equation}
and we shall refer to it as quasi-$T_2$.
The \acrshort*{CC2} Lagrangian is constructed based on the approximate
equations obtained for the singles and
doubles:\autocite{Christiansen1995-vy, Helgaker2000-tz}
\begin{equation}\label{eq:cc2-lagrangian}
  \begin{aligned}
  \mathcal{L}(\tamp{}, \tbar{})_\mathrm{CC2}
  &=
  E_0
  + \sum_{u=1}^{\mathcal{2}}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
  + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{Q_2} | \mathrm{HF}} \\
  &+ \braket{\tbar{1} |
  \check{\Phi} + \commutator{\check{\Phi}}{Q_2}
  | \mathrm{HF}}
  + \braket{\tbar{2} | \check{\Phi} | \mathrm{HF}}
  \end{aligned}
\end{equation}
The \acrshort*{CC2} methods brings the scaling of the \acrshort*{CCSD}
method down to $N^5$ in the iterative step.
From the approximate \acrshort*{CC2} Lagrangian we can derive the
multipliers equations:
\begin{subequations}\label{eq:cc2-multipliers}
  \begin{align}
    \denom{\mu_1}\tbar{\mu_1} &+
     \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}}
   + \braket{\tbar{1} |
       \commutator{\check{\Phi}}{\cluster{\mu_1}}
     + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{Q_2}
     | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} | \commutator{\check{\Phi}}{\cluster{\mu_1}} |
   \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tbar{\mu_2} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_2}} | \mathrm{HF}}
   + \braket{\tbar{1} |
       \commutator{\check{\Phi}}{\cluster{\mu_2}}
     | \mathrm{HF}}
    = 0
  \end{align}
\end{subequations}

The \gls{CC3} model is constructed analogously. Starting from the
\acrshort*{CCSDT} amplitudes equations in the $T_1$-transformed
representation, we retain only second order terms in the triples
equation:\autocite{Christiansen1995-pv, Koch1997-nm, Helgaker2000-tz,
Smith2005-an}
\begin{subequations}\label{eq:cc3-amplitudes}
  \begin{align}
    \denom{\mu_1}\tamp{\mu_1} &+ \braket{\mu_1 |
    \check{\Phi} + \commutator{\check{\Phi}}{T_2}
    + \commutator{\check{\Phi}}{Q_3}
    | \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tamp{\mu_2} &+ \braket{\mu_2 |
    \check{\Phi} + \commutator{\check{\Phi}}{T_2}
    + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2}
    + \commutator{\check{\Phi}}{Q_3}
    | \mathrm{HF}} = 0 \\
    \denom{\mu_3}\tamp{\mu_3} &+ \braket{\mu_3 |
    \commutator{\check{\Phi}}{T_2}
    | \mathrm{HF}} = 0
  \end{align}
\end{subequations}
In complete analogy with \acrshort*{CC2}, the quasi-$T_3$ operator was introduced:
\begin{equation}
  Q_3 = - \sum_{\mu_3}
  \denom{\mu_3}^{-1}\braket{\mu_3 | \commutator{\check{\Phi}}{T_2} | \mathrm{HF}}\cluster{\mu_3}.
\end{equation}
The \acrshort*{CC3} Lagrangian is constructed based on the approximate
amplitude equations
\eqref{eq:cc3-amplitudes}:\autocite{Christiansen1995-pv, Koch1997-nm,
Helgaker2000-tz, Smith2005-an}
\begin{equation}\label{eq:cc3-lagrangian}
  \begin{aligned}
  \mathcal{L}(\tamp{}, \tbar{})_\mathrm{CC3}
  &=
  E_0
  + \sum_{u=1}^{\mathcal{3}}\tbar{\mu_u}\denom{\mu_u}\tamp{\mu_u}
  + \braket{\mathrm{HF} | \check{\Phi} + \commutator{\check{\Phi}}{T_2} | \mathrm{HF}} \\
  &+ \braket{\tbar{1} |
  \check{\Phi} + \commutator{\check{\Phi}}{T_2} + \commutator{\check{\Phi}}{Q_3}
  | \mathrm{HF}} \\
  &+ \braket{\tbar{2} |
  \check{\Phi} + \commutator{\check{\Phi}}{T_2}
  + \frac{1}{2}\commutator{\commutator{\check{\Phi}}{T_2}}{T_2} + \commutator{\check{\Phi}}{Q_3}
  | \mathrm{HF}} \\
  &+ \braket{\tbar{3} | \commutator{\check{\Phi}}{T_2} | \mathrm{HF}}
  \end{aligned}
\end{equation}
The \acrshort*{CC3} methods brings the scaling of the \acrshort*{CCSDT}
method down to $N^7$ in the iterative step.
We can now derive the \acrshort*{CC3} multipliers equations:
\begin{subequations}\label{eq:cc3-multipliers}
  \begin{align}
    \denom{\mu_1}\tbar{\mu_1} &+
     \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_1}} | \mathrm{HF}}
   + \braket{\tbar{1} |
       \commutator{\check{\Phi}}{\cluster{\mu_1}}
     + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
     | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{2} |
       \commutator{\check{\Phi}}{\cluster{\mu_1}}
     + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
     + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{Q_3}
   | \mathrm{HF}} \nonumber \\
   &+ \braket{\tbar{3} |
      \commutator{\commutator{\check{\Phi}}{\cluster{\mu_1}}}{T_2}
     | \mathrm{HF}} = 0 \\
    \denom{\mu_2}\tbar{\mu_2} &+
    \braket{\mathrm{HF} | \commutator{\check{\Phi}}{\cluster{\mu_2}} | \mathrm{HF}}
   + \braket{\tbar{1} |
       \commutator{\check{\Phi}}{\cluster{\mu_2}}
     | \mathrm{HF}} \nonumber \\
  &+ \braket{\tbar{2} |
       \commutator{\check{\Phi}}{\cluster{\mu_2}}
     + \commutator{\commutator{\check{\Phi}}{\cluster{\mu_2}}}{T_2}
     | \mathrm{HF}} \nonumber \\
  &+ \braket{\tbar{3} |
       \commutator{\check{\Phi}}{\cluster{\mu_2}}
     | \mathrm{HF}}
    = 0 \\
    \denom{\mu_3}\tbar{\mu_3} &+
    \braket{\tbar{1} |
       \commutator{\check{\Phi}}{\cluster{\mu_3}}
     | \mathrm{HF}}
  + \braket{\tbar{2} |
       \commutator{\check{\Phi}}{\cluster{\mu_3}}
     | \mathrm{HF}}
    = 0
  \end{align}
\end{subequations}
